---
title: "Implicit Timecourse Project"
output:
  word_document:
    toc: yes
    toc_depth: 2
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 2
    number_sections: yes
    theme: united
  pdf_document:
    toc: yes
    toc_depth: '2'
    number_sections: yes
---

# Pre-amble

In this project we set out to characterize the speed of implicit adaptation under various circumstances. There are 4 experiments.

  1. Size of rotation + washout
  2. Feedback that is supposed to increase explicit adaptation (cursor-jump, terminal)
  3. Delays, before or after terminal feedback (before should be more explicit)
  4. A comparison with the time-course of re-aiming responses

In all conditions, 4 reach training targets were used, and after each training trial (with feedback) a no-cursor trial was completed. For these trials, participants were instructed to just move their hand straight to the target, regardless of how the cursor moved. After a brief familiarization task, participants did a 20+20 reach trial baseline, and a 100+100 reach trial adaptation phase. In the first experiment, this was followed by a 24+24 trial washout phase. In most conditions, there were 8 re-aiming responses given before the 57th, 61st, 65th, 69th, 73rd, 77th, 81st and 85th reach training trial. Except for the very last condition, where every reach training trial was preceded by a re-aiming response. (These were given by orienting an arrow in the screen, with 1 degree precision.)

Using the no-cursor reach aftereffects after every reach training trial, we can map out the speed of implicit adaptation with high temporal precision. The conditions are all supposed to evoke larger explicit strategies (or just overall adaptation, in the first experiment with rotation sizes). The re-aiming responses assess the magnitude of this strategy in each participant, such that we can also see if increased explicit strategies go hand in hand with decreased implicit adaptation (or vice versa) or whether or not these two kinds of adaptation are largely indepedent. Except for the last condition, we can't (and don't need to) assess the speed of explicit adaptation.

## Setting up R session

First we load all the code for this project:

```{r echo=T, results='hide', quietly=T, warn.conflicts=F}
source('R/data.R')
source('R/utilities.R')
source('R/figures.R')
source('R/statistics.R')
source('R/exponentials.R')
```

## Data

Before doing anything, we need to get the data on the local machine.

```{r eval=FALSE}
# download the raw data AND the processed data:
downloadData()
```

This downloads the relevant data from the associated [OSF repository](https://osf.io/ajwyr/).

## Processing

If we'd also want to redo all the processing, we could run the code in this cell:

```{r eval=FALSE}
# this summarizes each reach or re-aiming response to an angular deviation relative to the target
extractAngularDeviations()
```

For all reaches, this extracts the reach deviations at 0.2 times the home-target distance (i.e. 2 cm of a 10 cm reaching movement). And for aiming responses the aiming deviation from the target. The code to do so is provided, so if you think it should be done a different (and obviously better) way, the code can be changed / replaced.

You could even redo the bootstrapping of all the exponential fit parameters, even though it would take a long time. If you changed the way the dependent variables are calculated, this step can not be skipped though, as most of the statistics / conclusion depend on the bootstrapped exponential fits.

By default, this code uses 75% of available cores (the parallelization is meant to work on Linux and MacOS as well as on Windows) and does 5000 bootstraps for each fit. This could also be changed but it's not an easy, single function argument.

```{r eval=FALSE}
# these are sorted by how much time they take: most time at the top
bootstrapAllLearningExpFits()
bootstrapAllWashoutExpFits()
bootstrapAllAimingExpFits()
groupAvgFits()
```

The above three chunks of code have been set to not evaluate, but they can be run manually in the notebook.

# Experiment 1: rotation size

In the first part we gauge our method of interleaved no-cursor trials to see how this responds to 4 different rotation sizes (15, 30, 45, and 60 degree rotations). We don't find anything surprising. 

Here is a table about the sample of participants:

```{r}
knitr::kable(demographicsTable(exp=1))
```


Here is the plot of the data.

```{r fig.width=6, fig.height=9}
expBehaviorFig(exp=1, target='inline', timecoursemode = 'absolute')
```

## Table of behavioral descriptors

The table below lists group average estimates as well as bootstrapped 95% confidence intervals for several descriptors of behavior. The first 4 columns show parameters of exponential fits, for training reaches as well as no-cursor reaches. The first value is the optimal parameter for the average of the group data, the two values or the 2.5% and 97.5% value over 5k bootstraps. Then there are the re-aiming responses, which are simple the average and the upper and lower limit of a bootstrapped 95% confidence interval (1k values). The last two columns show exponential fits on the washout data, similar to the first 4 columns, but for simplicity we only show the starting points (here "asymptotes", need to re-label this in the paper) as the rates of changes are all very high (> 90%) and don't show any effects.

```{r}
knitr::kable(expTable(exp=1))
```

## First impression of figure and table

First, we confirm that with rotations smaller than 30 degrees, implicit reach aftereffects develop slower, while at all other rotation sizes, the rate of change is comparable, but only if expressed in real degrees change per trial, rather than percentage of asymptote. This casts some doubt on the use of rates/proportion to express speed of learning. Here we plot the exponential fits in real degrees, rather than relative percentages, but to allow comparisons with previous papers (like [Ruttle et al., 2021](https://doi.org/10.1038/s41598-021-81031-y) and [Ruttle et al., 2022](https://doi.org/10.1152/jn.00082.2022)) we also provide the group average rate of change as a percentage of asymptote. The statistics here are all done on the absolute degree change: which is the relative rate of change multiplied by the asymptote (which is roughly the degree change in the first trial, but informed by the full timecourse, so much less noisy).

For total adaptation, the two larger rotations behave similarly, with the rate of change decreasing for smaller rotations. The asymptotes for the two larger rotations seems similar as well, but perhaps the adaptation period was too short.

However, the washout periods show the same pattern for both regular reaches and no-cursor reaches. In the 2 largest rotations - if not the 3 largest rotations, there seems a significant portion of labile (implicit) adaptation. Starting with the first aligned trial in the washout phase, the 3 larger rotations are indistinguishable from each other. The 15 degree rotation shows smaller reach deviations in both types of trials, which signifies a qualitative difference between this small rotation and the 3 larger ones.

When looking at the relationship between explicit (re-aiming) and implicit (no-cursors), there is no linear relationship, although both seem to get a bit larger with increasing rotation size. Additionally, the distribution of our measures of implicit shows the same peak for the 3 larger conditions, and a smaller one for the smallest rotation; the same pattern as in the no-cursor time-courses and in the reach and no-cursor washouts. For explicit re-aiming strategies, we see uni-modal distributions around zero for the two smaller rotations, and bi-modal distributions for the two larger rotations, with 1 peak around zero and another peak at around 20 degrees. Maybe there is even a 3rd peak at 45 degrees for the largest rotation.

## Statistics

### Exponential fits

Here are the Bayesian F-tests on rate of change and asymptote of reach training:

```{r}
expFitBayesianFtest(exp=1,type='reaches')
```

Since both show effects, we look at between-group comparisons for successive rotation sizes:

```{r}
expFitGroupBayesianTtest(exp=1, type='reaches', mode='learning', timecoursemode='absolute', lambda=FALSE)
```

Here are the Bayesian F-tests on rate of change and asymptote of no-cursors:

```{r}
expFitBayesianFtest(exp=1, type='nocursors')
```

Again, we look at between-group comparisons for successive rotation sizes, and only for asymptote / N0:

```{r}
expFitGroupBayesianTtest(exp=1, type='nocursors', mode='learning', lambda=FALSE)
```

Here are the Bayesian F-tests on rate of change and asymptote of reach washout:

```{r}
expFitBayesianFtest(exp=1,type='reaches',mode='washout')
```

Again, we look at between-group comparisons for successive rotation sizes, but now also for lambda / RofC:

```{r}
expFitGroupBayesianTtest(exp=1, type='reaches', mode='washout')
```

Here are the Bayesian F-tests on rate of change and asymptote of no-cursor washout:

```{r}
expFitBayesianFtest(exp=1,type='nocursors',mode='washout')
```

In the figure it really does look like the 15 deg group has a lower N0 than the other groups. More or less like in reach training. So, despite a lack of an effect on the starting value (N0 is not the asymptote in this case) in washouts for no-cursors, we do look at between-group comparisons for successive rotation sizes:

```{r}
expFitGroupBayesianTtest(exp=1, type='nocursors', mode='washout', lambda=FALSE)
```

There might be some kind of effect comparing 15 to 45 degrees, but it's not very convincing.

### Aiming

We have complemented the continuous measures of implicit reach aftereffects with 8 re-aiming trials after 56 trials of training. Each set of 4 training and 4 no-cursor trials is preceded by a single re-aiming task, where people orient an arrow to indicate where they would move their hand in order to get the cursor on the target. The re-aiming trials are timed such that aiming does not affect the normal adaptation processes, as we have seen before ('t Hart et al., 2023, [bioRxiv](https://www.biorxiv.org/content/10.1101/2022.06.07.495044v3)). We expect little explicit strategy in the 15 and 30 degree conditions, and will see how this develops in larger rotations. Finally, we will also test if implicit adaptation can be predicted from explicit adaptation.

For those interested; in the last experiment we compare a continuous aiming condition with a control condition.

First we test if there is any effect of condition / rotation size on the magnitude of re-aiming responses:

```{r}
aimingBayesianFtest(exp=1)
```

There is, so that we now first compare the re-aiming responses to 0 for each condition, and then compare them between successive rotations.

Here are the comparisons with 0:

```{r}
aimingZero()
```

The evidence goes toward some small amount of re-aiming in the 15 degree condition, although this is largely driven by 2 participants and it does not reach threshold (BF=3). There is some re-aiming happening in the 30 degree condition. In Fig 2XXX we can see that the distribution of re-aiming responses is still close to zero, but it is about 3 degrees. In the other two conditions, it is much more clear that almost all participants engage in some amount of re-aiming. This can both be seen in Fig 2XXX and is shown by the statistics above.

Now we compare aiming between successive rotations.

```{r}
aimingGroupTtest(exp=1)
```

Despite there being pretty good evidence for re-aiming strategies in the 30 degree group and not in the 15 degree group, re-aiming is not very different between the two groups. We do see the expected different between the 30 and 45 degree group, but it is only approaching significance when comparing the 45 anf 60 degree group. Perhaps this can be explained by the larger spread with larger rotations. 

### Linear prediction of measures of implicit from measures of explicit adaptation

Here we test explicit adaptation as a predictor of implicit adaptation.

```{r}
testGroupLinearAdditivity(exp=1)
```

There is a significant linear relationship in the 60 degree group only, but this is far from linear. In the groups with lower rotations, the relationship is not even significant.

## Conclusions

The interleaved trial method works here as well. The explicit re-aiming trials also behave more or less as expected. To test how different kinds of feedback impact the time course of implicit adaptation - in particular in the presence of explicit adaptation as well, we will continue using the 45 degree rotation. We will skip the washout phases however, as they don't seem to add much to other measures. This also leaves time for additional measures.

There are no effects on rates of change whatsoever - if they are expressed in degrees reach deviation, rather than as a percentage of asymptote.

Reach training aymptotes keep increasing with rotation size.

Similarly for training no-cursors, except that they seem maxed out for 45 degrees. There is even a small difference between 30 and 45 degrees.

For washout reaches, the starting point for the 15 degrees group is lower than the other three groups. Unlike during training, no-cursors during washout don't show a clear pattern. It may be that the 15 degree group has somewhat lower starting point compared to some groups, but the evidence is inconclusive.


# Experiment 2: feedback types

Here we test the timecourse of implicit adaptation, with two different kinds of feedback. Previously, the cursor was visible continuously, and was rotated right from the start. Here, in one condition, we keep the cursor aligned the first 50% of the reach, and introduce the rotation only after 50% of the reahc is completed. At that point, the cursor jumps from aligned to 45 degrees rotated: the _"cursor-jump"_ condition. Then, we also introduce a condition where the cursor is not shown during the reach, or during the return. It is only shown at 1 point: the first point that is at or beyond the distance of the target. The feedback is shown is for XXX ms. (During the return to the start position, different feedback is provided to allow returning to the start position without given position feedback.) We call this the _"terminal"_ condition. Both are supposed to increase explicit learning, and should or could reduce implicit learning. This has been shown before, but the time-course of implicit learning has not been characterized yet (but see Ruttle et al., 2022). Apart from these 2 new conditions, we collect more participants doing a 45 degree rotation (albeit without the washout) and we combine these participants with the existing 45 degree as a reference condition for all new experiments, and we call this the _"control"_ condition.

Here is a table about the sample of participants:

```{r}
knitr::kable(demographicsTable(exp=2))
```

Here is the data:

```{r fig.width=6, fig.height=7.5}
expBehaviorFig(exp=2, target='inline', timecoursemode = 'absolute')
```

## Table

The table below is similar to that for experiment 1, except that there are no washout phases.

```{r}
knitr::kable(expTable(exp=2))
```

## Impressions left by figure and table

Overall adaptation seems equal for the three groups, but the control group appears to have larger implicit aftereffects at the end. The timecourses don't show much of a difference between the groups for overall adaptation. However, the implicit, no-cursor timecourses do look qualitatively different. Implicit adaptation seems suppressed for cursor-jump and more so for terminal feedback, in terms of the rate of change (in degrees) per trial.

The distribution of explicit (re-aiming) strategies, is no longer clearly bi-modal with 1 peak at 0 explicit strategies for the two new conditions. It would seem that _all_ participants developed some kind fo strategy in these conditions. The distribution of implicit (no-cursors) seems to have changed in the control group as well, because of the addition of extra participants. Where before there was only the impression of a long-tailed distribution, now it seems that there are two distinct peaks. The conditions are exactly identical however, so we have no reason to interpret this as anything else but natural variance. As also seen in the timecourse, implicit adaptation was larger in the control group than in the other groups. The asymptotic levels of implicit adaptation are suppressed, but they do not disappear. Finally, there is no clear relationship between these measures of implicit and explicit components of adaptation.

## Statistics

Here are the Bayesian F-tests on rate of change and asymptote of reach training:

```{r}
expFitBayesianFtest(exp=2,type='reaches')
```

We're not doing any follow-ups because there are no effects, and that's what it looks like in the figure as well.

Here are the Bayesian F-tests on rate of change and asymptote of no-cursors:

```{r}
expFitBayesianFtest(exp=2, type='nocursors')
```

It _looks_ like the implicit time courses develop at different speeds in the figure, so we'll test those, despite there not being any main effect.

```{r}
expFitGroupBayesianTtest(exp=2, type='nocursors', mode='learning', timecoursemode='absolute', N0=FALSE)
```

But there really is no effect...

### Aiming

Testing for an effect of condition / group on re-aiming:

```{r}
aimingBayesianFtest(exp=2)
```

Seems there is an effect, so now we compare the control group to the others:

```{r}
aimingGroupTtest(exp=2)
```

Both groups are different from control.

### Linear prediction of implicit from explicit

Here we test explicit adaptation as a predictor of implicit adaptation.

```{r}
testGroupLinearAdditivity(exp=2)
```

None of the relations are significant, and confidence intervals of the slopes include 0 but not -1.

## Conclusions

These two feedback types indeed increased explicit adaptation, which may have led to decreased implicit adaptation, but the relationship is not 1:1.

# Experiment 3: delays with terminal feedback

The terminal feedback in the previous experiment should lead to "explicit-only" adaptation, if there is a delay after reach completion and before the feedback is shown. We test this in a variant on the previous "terminal" condition where the feedback is presented after a delay, the _"delay -> feedback"_ condition. This changes both the overall duration of the experiment, as well as the time between the onset of trials. This could also change learning, so we matched this with a condition where the terminal feedback is shown first, and then there is a delay of the same duration. We call this the _"feedback -> delay"_ condition.

Here is a table about the sample of participants:

```{r}
knitr::kable(demographicsTable(exp=3))
```

The data is shown here:


```{r fig.width=6, fig.height=7.5}
expBehaviorFig(exp=3, target='inline', timecoursemode = 'absolute')
```

## Table

```{r}
knitr::kable(expTable(exp=3))
```

## First observations

In the two new conditions, overall adaptation is comparable in both speed and asymptote with the previous terminal condition, as well as the control condition. The speed of implicit adaptation seems to be in between that of the previous terminal condition and the control condition. In both conditions with delay there is a wide range of explicit re-aiming strategies, including 0 strategy this time. Distributions of implicit adaptation are comparable in all 3 terminal feedback condition: the presence of delays, or the order of feedback and delay do not seem to affect the asymptotic levels of implicit adaptation. Given the consistently low levels of implicit adaptation, it should not be surprising there was no relationship between explicit and implicit measures of adaptation.

## Statistics

Here are the Bayesian F-tests on rate of change and asymptote of reach training:

```{r}
expFitBayesianFtest(exp=3,type='reaches')
```

These Bayes Factors indicate both the rates of change and the asymptotes in overall reach training are equal between all groups. We won't do any follow-ups.

Here are the Bayesian F-tests on rate of change and asymptote of no-cursors:

```{r}
expFitBayesianFtest(exp=3, type='nocursors')
```

This also does not indicate any differences.

### Aiming

We see if there is any effect on aiming responses:

```{r}
aimingBayesianFtest(exp=3)
```

There is an effect. Slightly surprising as there are no effects on overall learning or implicit reach aftereffects. Let's see where the differences are.

```{r}
aimingGroupTtest(exp=3)
```

It appears that in the terminal feedback without any kind of delay, the re-aiming responses are larger than in all other groups. That is: added delays do not seem to make learning more explicit in this data set. Perhaps we want to see if the delay conditions are different from control as well.

### Linear correlation between implicit and explicit adaptation

Here we test explicit adaptation as a predictor of implicit adaptation.

```{r}
testGroupLinearAdditivity(exp=3)
```

The two untested groups here (the two with delays) again show no linear relationship between implicit and explicit adaptation.

## Conclusion

Terminal feedback did increase explicit adaptation, and decreased implicit adaptation. While the RofC of implicit adaptation might counterintuitively be slighty larger again with added delays, the delays had no other effects: asymptotic levels of explicit and implicit adaptation remained largely the same as those in the condition with terminal feedback without delays.

# Experiment 4: continuous aiming

So far, we have only used 8 re-aiming trials when adaptation was relatively close to saturation. In this experiment, we want to see if adding a re-aiming trial before every trial affects explicit adaptation, and perhaps implicit adaptation as well. So we added a condition where every reach training trial was preceded by a re-aiming assessment (arrow orienting). We call this the _"aiming"_ condition.

Here is a table about the sample of participants:

```{r}
knitr::kable(demographicsTable(exp=4))
```

Here is the data:

```{r fig.width=6, fig.height=7.5}
expBehaviorFig(exp=4, target='inline', timecoursemode = 'absolute')
```
## Table

```{r}
knitr::kable(expTable(exp=4))
```

## First observations

Magenta is aiming (should be added to the legend somewhere...). The asymptote for the continuous aiming group looks a bit higher, at least at the start. We can see this in the exponential timecourse plots (fit on only the data up to where the first aiming is seen in most conditions: the 57th trial). At the end of the 100 trials of adaptation, levels seem more similar again, but it might just been that having to think about strategies in terms of directions of hand movements, gives away information to participants on the nature of the perturbation, and perhaps how to deal with it - acting as a partial instruction on how to counter the rotation. The no-cursor timecourse is similar in the continuous aiming group and the control group. The timecourse of explicit adaptation rises much faster though. The level of implicit adaptation in the new continuous adaptation group seems uni-modal, aligning with the first component in the distribution of implicit adaptation in the control group, so that it's overall slightly smaller. The distribution of explicit adaptation also looks uni-modal, but it's on the higher end compared to the control group. That is: for 45 degree rotations, the introduction of continuous aiming does not change behavior as much as with 30 degree rotations ('t Hart et al., under review).

## Statistics

Here are the Bayesian F-tests on rate of change and asymptote of reach training:

```{r}
expFitBayesianFtest(exp=4,type='reaches')
```

Here are the Bayesian F-tests on rate of change and asymptote of no-cursors:

```{r}
expFitBayesianFtest(exp=4, type='nocursors')
```

Adding continuous aiming does not seems to affect the speed or extent of overall learning or implicit adaptation.

### Aiming

Let's see if there is a difference in aiming between the continuous aiming group and the control group.

```{r}
aimingGroupTtest(exp=4)
```

Looking at the data in the figure, the re-aiming responses in the continuous aiming group turned out higher than those in the control group.

### Implicit-explicit correlation

Here we test explicit adaptation as a predictor of implicit adaptation.

```{r}
testGroupLinearAdditivity(exp=4)
```

Also in the aiming group, there seems to be no relationship between measures of implicit and measures of explicit adaptation.

## Conclusion

For rotations that are already likely to elicit more of an explicit strategy, the introduction of aiming assessments before every reach training trials does change behavior, but not as much as we've seen for smaller rotations.

# Export manuscript figures

```{r eval=F}
targets <- c('pdf') # svg, png, tiff
for (exp in c(1:4)) {
  for (target in targets) {
    expBehaviorFig(exp=exp, target=target)
  }
}
```

